Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-12-03 tmm.tex
The submittable version

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and provide the user with best online video watching experience, an adaptive video streaming system which can adjust video quality according to actual network condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. Then on that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm can significantly reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed system integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 1.24 to 0.69, and at the same time deliver higher video quality, with an average PSNR gain of 0.83dB. The system is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios include online Video-on-Demand (VoD), live broadcasting, video-based online education, etc. In these applications, video data is constantly delivered to different clients over various transmission channels, and the client media player can begin playing the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch video anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community during the past few years \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and becomes an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network condition significantly influences the user's watching experience. To provide better network condition, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm named OpenFlow, noticeable results are achieved, e.g., a solution for improving network throughput in hybrid network \cite{Li14} and a demonstration of OpenFlow-controlled network orchestration for video manycast \cite{Xue15}. Although these studies have offered a promising alternative, currently traditional best-effort networks still dominate, and the variety of users' network condition remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to Cellular). Even in the same network, the end-to-end bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable, and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a constant bitrate. To cope with this challenge, the video streaming system should be able to automatically adjust the video's bitrate or quality according to the network condition. In other words, the video streaming system needs to be adaptive.

@@

Due to the easy deployment of simulcast systems \cite{Bouten14}, they are already used by many commercial video websites (e.g., YouTube \cite{YouTube}) to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams require a lot of computation time and disk space. In order to provide a wide range of seamless quality adaptation at reasonable time and space cost, the adoption of SVC is necessary. An SVC video bitstream can be extracted at the fine granularity of one NALU (a very small partition of data), and performance analysis \cite{SVCPerformance} shows that SVC clearly outperforms simulcast coding in terms of compression efficiency. Therefore, to adopt SVC in the video streaming framework is technically a better option and has attracted much attention in the study \cite{Chuah12}-\cite{Cicalo14}.

@@

The following of this paper will present our solution for each of the problems. The solutions then work together and serve as a solid foundation for a practical adaptive video streaming system. It should be noted that, in practical streaming systems there are other issues to be addressed. One of them is occasional packet loss during transmission, which will cause decoding errors and corrupted images, resulting in an significant degradation of video quality. How to avoid or even recover from packet loss/error is an important research topic that is popular in both networking and multimedia community. However, in this paper which focuses on server-side adaptation, packet loss is not handled and the video stream is assumed to be successfully delivered to the client.

@@

Obviously, $e_j$ is only determined by packet $j$, and has no relation with other enhancement packets. The value $e_j$ is defined as ``error vector'' of packet $j$, which represents the pixel value error caused by discarding the packet $j$. We can obtain the error vector of each enhancement packet by subtraction between video sequences reconstructed from two enhancement packet subsets $I_1$ and $I_2$, with $I_1 \subseteq I_2$ and $I_2 \setminus I_1 =\{j\} $. Once all enhancement packets' error vectors are obtained, the distortion caused by discarding a set of enhancement packets can be estimated by adding up the error vector of each packet in the set. For example, if enhancement packets $k$, $l$, $n$ are removed from a bitstream, then the difference of the reconstruction before and after the removal is supposed to be $e = e_k + e_l + e_m$. The reason for this sum to be rational is twofold: first, unlike MSE or PSNR, error vector is simply the difference of raw pixel values, with no quadratic calculation ever involved; second, each enhancement packet adds its own refinement to the pixel values during the reconstruction, due to the linear feature of the decoding process (see the derivation above). The second point is conditionally true though, as noted below.

Note that in equations (\ref{eq:base0})-(\ref{eq:subset2}) the value of ${\rm\bf M}$ and $p$ remain the same, since in the case of quality scalability the enhancement packets are generated only by manipulating the transform coefficients, and not interfered with the motion compensation and static prediction. This argument no longer holds for the other scalable dimensions (e.g., spatial scalability), so LEM does not simply apply for those cases. Therefore, only quality scalable videos are used in our system, and further research for other scalable dimensions can be our future work.

@@

Consider a video bitstream with 8-frame GOPs shown in Fig. \ref{fig:GOP_Structure}. Base layer packets are colored black, and enhancement packets with the same color are allocated in one group. In each group, the enhancement packets' error vectors can be calculated simultaneously, since the pixel errors they can cause are not interfered with each other. For example, the removal of packet 8 would cause pixel errors only in frame 1, 2 and 3, while packet 10 only in frame 5, 6 and 7 (frame index starts from 0). So we can do subtraction only once for this group, to obtain error vectors $e_{8}$ and $e_{10}$, for both packets 8 and 10, as well as peer packets in all other GOPs. Note that these error vectors are highly sparse ones, because the discarding of a single enhancement packet can bring distortion only to a limited number of frames. To acquire all error vectors, the number of required extraction and decoding is approximately the number of enhancement layers $(L_Q - 1)$ times the temporal layer number $L_T$. For the example in Fig. \ref{fig:GOP_Structure}, $L_Q = 3$ and $L_T = 4$, so the extraction-and-decoding number is $(3 - 1) \times 4 = 8$. However, we have to do the same manipulation for $L_Q - 1 = 2$ more times, to dissociate the distortion impact of ``key frame'' \cite{H264Overview} (e.g. frame 0, 8, 16, etc.) packets. The loss of one enhancement packet in a key frame would bring distortion to two neighboring GOPs. As an instance, the loss of packet 4 would propagate error into frame 1 to 15, which would interfere with the distortion impact of packets in frame 0 and 16 (packet 1, 2 and 19, 20). So key frame enhancement packets within the same quality layer (e.g. packet 2, 4 and 20) should be allocated in two groups -- one for key frame packets of even number GOPs (packet 2 and 20), the other for those of odd number GOPs (packet 4). Therefore, the actual number of extraction and decoding for error vector acquisition is $(L_Q - 1) \cdot (L_T + 1)$, approximately the same with the Quality Layer information acquisition in JSVM \cite{Amonou07}. Since computation is mainly consumed with this decoding in the bitstream extraction algorithm, the computation complexity has no obvious promotion compared with JSVM.

@@

Before using the estimation based on LEM, we conduct an experiment to verify it. All bitstreams in the experiment are encoded with the JSVM encoder, where the GOP structure as illustrated in Fig. \ref{fig:GOP_Structure} and the CABAC entropy coding are used. Each bitstream contains one base layer and one enhancement layer, with QP = 33 and QP = 27, respectively. The enhancement layer is further divided into 2 Medium Grain Scalability (MGS) layers by configuring the MGS vector to separate the transform coefficients into two groups (containing 4 and 12 coefficients respectively). Random combinations of enhancement packets are removed from a bitstream, and the distortion estimated by the linear error model are compared with the actually calculated distortion. Decoding is performed by JSVM decoder, and distortion estimation or calculation is done by Matlab. One comparison result for the test sequence {\em Foreman} (only first 33 frames are used) is shown in Fig. \ref{fig:model_verification}. The horizontal axis stands for frame index, and the vertical axis for MSE measurement. The solid line represents the actual distortion for the frames, and the dot line represents the distortion estimated by the proposed linear error model. It can be seen that the two lines are very close to each other, which demonstrates the accuracy of estimation using the proposed model. Estimation errors for all the eight SVC standard test sequences are listed in Table \ref{tab:estimation-error}, which shows that the average error is only about 5\%.

@@

In this subsection, we utilize the linear error model to enable R-D optimized bitstream extraction. Since finding the global optimal solution of the bitstream extraction problem is impractical, our bitstream extraction algorithm uses a greedy method \cite{GreedyAlgo} to get a sub-optimal solution. Note that the greedy algorithm is able to find a local optimum. However, the gap between the local optimum and the global optimum is uncertain. Similar to the greedy solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtraction of distortions before and after the packet is discarded.

@@

For long video sequences, an optimization window containing a certain number of frames is usually needed, and the priority assignment and bitstream extraction is conducted independently in each window. Suppose the size of the optimization window is $K$, i.e., there are $K$ frames involved in the above process, and each frame contains $L_Q$ quality layers (one base layer and $L_Q-1$ enhancement layers). Then the total number of enhancement packets that can be discarded is $K \cdot (L_Q-1)$, and in every greedy step, there are at most $K$ packets, i.e., the packets currently at the top of every frame (see analysis in \ref{subsec:analysis-extraction}), that can be selected to discard. Given that, the priority assignment process in each optimization window can be described in Algorithm \ref{algo:greedy}, with some terms in it explained below. The initial value $e_{full}$ is the error between the video sequence decoded from full enhancement packets and the original video sequence, as already introduced in (\ref{eq:subset_error}). $\Phi_m$ is defined as:
\begin{equation}
\label{eq:R-D_impact_m}
\Phi_m = \min_{i \in I_{top}} \Phi_i \: ,
\end{equation}
where $I_{top}$ stands for the top-layer packets, those currently able to be discarded, and $\Phi_i$ denotes the RD impact of packet $i$. $\Phi_i$ is calculated based on (\ref{eq:rd_impact}):
\begin{equation}
\label{eq:R-D_impact_i}
\Phi_i = \dfrac{\partial D}{\partial R} = \dfrac{PSNR(e_{seq}) - PSNR(e_{seq} + e_i)}{SIZE(i)} \:,
\end{equation}
where $SIZE(i)$ is the size of packet $i$, and $PSNR(*)$ is the function to obtain PSNR from an error vector. To be clear, PSNR is defined via mean squared error (MSE), i.e., the average of the squares of the elements in the error vector.

\begin{algorithm}
\caption{Greedy-like priority assignment algorithm}
\label{algo:greedy}
\begin{algorithmic}
    \STATE Initialize $e_{seq}$ to $e_{full}$, and $q$ to 0.
    \WHILE{$q < K \cdot (L_Q - 1)$}
	    \FOR{the (at most) $K$ top layer packets}
		    \STATE Find the packet $m$ which has the least RD impact $\Phi_m$
		\ENDFOR
		\STATE Set the priority value of packet $m$ to $q$
		\STATE Remove $m$ from the sequence
		\STATE Update $e_{seq}$ to $e_{seq} + e_m$, and update $q$ to $q+1$
    \ENDWHILE
\end{algorithmic}
\end{algorithm}

It is easy to see that the computational complexity of this priority assignment algorithm is $O(K^2)$. Note that, however, the computation in the priority assignment process is actually very insignificant compared with the decoding process needed in the error vector acquisition. A larger optimization window means we have more packets to compare and determine their priorities, which will potentially result in more optimized extraction. However, doing this will also increase the computational complexity and, more noticeably, increase the memory used to store the error vectors. Besides, if the priority values are to be stored and signalled using the 6-bit priority identifier field of the NALU header \cite{Amonou07}, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

In some scenarios, the priority assignment and bitstream extraction needs to be conducted without access to the original video sequence. This would make the bitstream extraction a more difficult problem, since the distortion is harder to obtain for lacking of reference signal. The ``no-reference bitstream extraction'' problem is actually more frequently encountered in real practice, since it is common that streaming service provider only have the compressed bitstream rather than the video source. In this subsection, we consider this no-reference case and modify the proposed algorithm to meet the special characteristic caused by lacking of reference.

Without the original video sequence, we can not get the error between the video sequence decoded from full enhancement packets and the original sequence, which is $e_{full}$ used in \ref{algo:greedy}. A straightforward idea is to set $e_{full}$ to zero, meaning all distortions are calculated with respect to the fully reconstructed signal, rather than to the original sequence. However, simply doing so causes the algorithm to perform badly in experiments. It is found that, since discarding some enhancement packets will not bring much significant error, the calculated PSNR values are too large to well represent the real distortion. To address this issue, instead of calculating PSNR, we directly use MSE of all pixels as the distortion criteria in the priority assignment algorithm. This proves to be an effective solution, and the extraction performance turns out to be much better.

@@

PID controller \cite{PID}\cite{Astrom02} is a control loop feedback mechanism commonly used in industrial control systems. \textit{Process variable} and \textit{setpoint} are two core concepts in PID control. The process variable stands for the current status of the system. The setpoint is the target for this control system to reach. The goal of PID control is to minimize the error between process variable and setpoint by adjusting some control variable. The controller output $u(t)$ is defined as follows:
\begin{equation}
\label{eq:pid-output}
u(t) = {K_p} \cdot e(t) + {K_i} \cdot \int_0^t {e(\tau )d\tau }  + {K_d} \cdot \frac{d}{{dt}}e(t) \: ,
\end{equation}
where $e(t)$ represents the error between process variable and setpoint at time $t$. $K_p$, $K_i$, and $K_d$ here are tuning parameters representing the proportional gain, integral gain and derivative gain, respectively. The result of (\ref{eq:pid-output}) is used for manipulating the system to make process variable change towards setpoint.

As an effective and flexible control technique, PID control has been used to solve all kinds of control problems. In the area of video coding, we notice that PID is utilized to design rate control algorithms (rate control of video coding is mainly to adjust the encoding parameters for every frame, so that the bits are appropriately allocated to avoid buffer overflow and achieve desired bitrate) \cite{Wong04}\cite{Yang10}. We also notice that some researchers have proposed theoretical model or middle-ware framework for QoS adaptation \cite{Li98}\cite{Li99}, with PID as the basic control scheme. However, we have not found any paper that integrates PID method into the quality control of video streaming. While rate control of video coding and quality control of video streaming are two different problems, they do share some common properties, such as the buffer mechanism and the requirement for smoothness. That inspires us to design a PID-based quality control algorithm and utilize it to deliver better video quality to the user.

@@

In DSS, two timelines are important. Push timeline indicates the time of pushing video data packets to streaming buffer, and play timeline indicates the time these data packet should be played, i.e., the presentation time stamp (often seen as PTS) of each packet. Their relationship is shown in Fig. \ref{fig:intervals}. The program will check the timelines periodically, and, in the default quality control algorithm of DSS, use the packet delays to decide whether quality change is needed. The two intervals shown in Fig. \ref{fig:intervals}, i.e., check interval and actual interval, are worth notice and may provide useful information. In the ideal situation, during each check interval, the server pushes an amount of data that will actually play for a period of time equal to that check interval, i.e., \textit{actual\_interval} = \textit{check\_interval}. If \textit{actual\_interval} is smaller than \textit{check\_interval}, it means that less data is pushed to the client, and the pushing is slowed down because bandwidth is relatively low at this time. On the other hand, if \textit{actual\_interval} is larger than \textit{check\_interval}, it indicates the network condition is good and data is being pushed quickly to the client. The ratio of these two intervals, i.e., \textit{actual\_check interval ratio}, is in direct proportion to the ratio of the bitrate suitable for the network condition and the bitrate currently being used.

@@

This ratio-based model is more simple and effective for the chosen process variable and setpoint. Moreover, in this model the controller output has intuitive meaning when the three parameters are normalized. It is easy to know that all of the three PID parts ($E_p^t$, $E_i^t$, $E_d^t$) are ratios with a value more or less than 1, and the controller output $u_t$ can be seen as a weighted average of the three ratios, with the three parameters as weights. Therefore the controller output is also a ratio value around 1, which reflects the relationship between the bitrate suitable for the network condition and the bitrate currently being used. In other words, if the output is larger than 1, then the controller has determined the bitrate should be increased accordingly, and vice versa. Use it as a multiplier to manipulate the bitrate currently being used, and we would be able to control the streaming system to match the network condition.

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and improve the experience of watching videos online, an adaptive video streaming system which can adjust video quality according to actual network condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. On that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm is able to reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed system integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 1.24 to 0.69, and at the same time deliver higher video quality, with the PSNR average increased by 0.83 dB. The system is deployed at online video website www.7dlive.com and proves to be practically effective.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios include Video-on-Demand (VoD), live broadcasting, video-based online education, etc. In these applications, video data is constantly delivered to different clients over various transmission channels, and the client media player can begin to play the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch video anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community during the past few years \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and becomes an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network condition significantly influences the user's watching experience. To provide better network condition, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm named OpenFlow, noticeable results are achieved. For instance, a solution for improving network throughput in hybrid network is proposed in \cite{Li14} and a demonstration of OpenFlow-controlled network orchestration for video manycast is present in \cite{Xue15}. Although these studies have offered a promising way to optimize network condition, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to Cellular). Even in the same network, the end-to-end bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable, and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a fixed bitrate. To cope with this challenge, the video streaming system should be able to adjust the video's bitrate or quality according to the network condition. In other words, the video streaming system needs to be adaptive.

@@

Due to the easy deployment of simulcast systems \cite{Bouten14}, they are already used by many commercial video websites (e.g., YouTube \cite{YouTube}) to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams require a lot of computation time and disk space. In order to provide a wide range of seamless quality adaptation at reasonable time and space cost, the adoption of SVC is necessary. An SVC video bitstream can be extracted at the fine granularity of one NALU (a very small partition of data), and performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast where the same content is encoded multiple times. Therefore, to adopt SVC in the video streaming framework is technically a better option and has attracted much attention in the academia \cite{Chuah12}-\cite{Cicalo14}.

@@

The following of this paper will present our solution for each of the problems. The solutions then work together and serve as a solid foundation for a practical adaptive video streaming system. It should be noted that, in practical streaming systems another much concerned issue is the occasional packet loss during transmission, which will cause decoding errors and corrupted images, resulting in an significant degradation of video quality. How to avoid or even recover from packet loss/error is an important research topic that is popular in both networking and multimedia community. However, in this paper which focuses on server-side adaptation, packet loss is not handled and the video stream is assumed to be successfully delivered to the client.

@@

Obviously, $e_j$ is only determined by packet $j$, and has no relation with other enhancement packets. The value $e_j$ is defined as ``error vector'' of packet $j$, which represents the pixel value error caused by discarding the packet $j$. We can obtain the error vector of each enhancement packet by subtraction between video sequences reconstructed from two enhancement packet subsets $I_1$ and $I_2$, with $I_1 \subseteq I_2$ and $I_2 \setminus I_1 =\{j\} $. Once all enhancement packets' error vectors are obtained, the distortion caused by discarding a set of enhancement packets can be estimated by adding up the error vector of each packet in the set. For example, if enhancement packets $x$, $y$, $z$ are removed from a bitstream, then the difference of the reconstruction before and after the removal is supposed to be $e = e_x + e_y + e_z$. The reason for this sum to be rational is twofold: first, unlike MSE or PSNR, error vector is simply the difference of raw pixel values, with no quadratic calculation ever involved; second, each enhancement packet linearly adds its own refinement to the pixel values during the reconstruction, due to the linear feature of the decoding process (see the derivation above). It is conditional for the second point to be true though, as noted below.

Note that in equations (\ref{eq:base0})-(\ref{eq:subset2}), the reason why the values of ${\rm\bf M}$ and $p$ remain the same is that, in the case of quality scalability the enhancement packets are generated only by manipulating the transform coefficients, and not interfered with the motion compensation and static prediction. This argument no longer holds for the other scalable dimensions (e.g., spatial scalability, where the video resolution is changed), so LEM does not simply apply for those cases. Therefore, only quality scalable videos are used in our system, and further research for other scalable dimensions can be our future work.

@@

Consider a video bitstream with 8-frame GOPs shown in Fig. \ref{fig:GOP_Structure}. Base layer packets are colored black, and enhancement packets with the same color are allocated in one group. In each group, the enhancement packets' error vectors can be calculated simultaneously, since the pixel errors they can cause are not interfered with each other. For example, the removal of packet 8 would cause pixel errors only in frame 1, 2 and 3, while packet 10 only in frame 5, 6 and 7 (frame index starts from 0). So we can do subtraction only once for this group, to obtain error vectors $e_{8}$ and $e_{10}$, for both packets 8 and 10, as well as peer packets in all other GOPs. Note that these error vectors are highly sparse ones, because the discarding of a single enhancement packet can bring distortion only to a limited number of frames. To acquire all error vectors, the number of required extraction and decoding is approximately the number of enhancement layers $(L_Q - 1)$ times the temporal layer number $L_T$. For the example in Fig. \ref{fig:GOP_Structure}, $L_Q = 3$ and $L_T = 4$, so the extraction-and-decoding number is $(3 - 1) \times 4 = 8$. However, we have to do the same manipulation for $L_Q - 1 = 2$ more times, to dissociate the distortion impact of ``key frame'' \cite{H264Overview} (e.g. frame 0, 8, 16, etc.) packets. The loss of one enhancement packet in a key frame would bring distortion to two neighboring GOPs. As an instance, the loss of packet 4 would propagate error into frame 1 to 15, which would interfere with the distortion impact of packets in frame 0 and 16 (packet 1, 2 and 19, 20). So key frame enhancement packets within the same quality layer (e.g. packet 2, 4 and 20) should be allocated in two groups -- one for key frame packets of even number GOPs (packet 2 and 20), the other for those of odd number GOPs (packet 4). Therefore, the actual number of extraction and decoding for error vector acquisition is $(L_Q - 1) \cdot (L_T + 1)$, approximately the same with the Quality Layer information acquisition in JSVM \cite{Amonou07}.

@@

Before using the estimation based on LEM, we conduct an experiment to verify its accuracy. Bitstreams in the experiment are encoded using the JSVM encoder, with the GOP structure illustrated in Fig. \ref{fig:GOP_Structure} and the CABAC entropy coding. Each bitstream contains one base layer and one enhancement layer, with QP = 33 and QP = 27, respectively. The enhancement layer is further divided into 2 Medium Grain Scalability (MGS) layers by configuring the MGS vector to separate the transform coefficients into two groups (containing 4 and 12 coefficients respectively). Random combinations of enhancement packets are removed from a bitstream, and the distortion estimated by the linear error model are compared with the distortion actually calculated after real decoding. Decoding is performed by JSVM decoder, while other operations are done by MATLAB. One comparison result for the test sequence {\em Foreman} (only first 33 frames are used) is shown in Fig. \ref{fig:model_verification}. The horizontal axis stands for frame index, and the vertical axis for MSE measurement. The solid line represents the actual distortion of the frames, and the dot line represents the distortion estimated by the proposed linear error model. It can be seen that the two lines are very close to each other, which demonstrates the accuracy of estimation using the proposed model. Estimation errors for all the eight SVC standard test sequences are listed in Table \ref{tab:estimation-error}, which shows that the average error is only about 5\%.

@@

In this subsection, we utilize the linear error model to enable R-D optimized bitstream extraction. Since finding the global optimal solution of the bitstream extraction problem is impractical, our bitstream extraction method uses a greedy algorithm \cite{GreedyAlgo} to get a sub-optimal solution. Note that the greedy algorithm is able to find a local optimum. However, the gap between the local optimum and the global optimum is uncertain. Similar to the greedy solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtraction of distortions before and after the packet is discarded.

@@

For long video sequences, an optimization window containing a certain number of frames is needed, and the priority assignment and bitstream extraction is conducted independently in each window. Suppose the size of the optimization window, i.e., the number of frames involved in the above process, is $K$, and each frame contains $L_Q$ quality layers (one base layer and $L_Q-1$ enhancement layers). Then the total number of enhancement packets to be discarded is $K \cdot (L_Q-1)$, and in every greedy step, there are at most $K$ packets, i.e., the packets currently at the top of every frame (see analysis in \ref{subsec:analysis-extraction}), that can be selected to discard. Given that, the priority assignment process in each optimization window can be described in Algorithm \ref{algo:greedy}, with some terms in it explained below. The initial value $e_{full}$ is the error between the video sequence decoded from full enhancement packets and the original video sequence, as already introduced in (\ref{eq:subset_error}). $\Phi_m$ is defined as:
\begin{equation}
\label{eq:R-D_impact_m}
\Phi_m = \min_{i \in I_{top}} \Phi_i \: ,
\end{equation}
where $I_{top}$ stands for the top-layer packets, those currently able to be discarded, and $\Phi_i$ denotes the RD impact of packet $i$. $\Phi_i$ is calculated based on (\ref{eq:rd_impact}):
\begin{equation}
\label{eq:R-D_impact_i}
\Phi_i = \dfrac{\partial D}{\partial R} = \dfrac{PSNR(e_{seq}) - PSNR(e_{seq} + e_i)}{SIZE(i)} \:,
\end{equation}
where $SIZE(i)$ is the size of packet $i$, and $PSNR(*)$ is the function to obtain PSNR from an error vector. To be clear, PSNR is defined via mean squared error (MSE), i.e., the average of the squares of the elements in the error vector.

\begin{algorithm}
\caption{Greedy-like priority assignment algorithm}
\label{algo:greedy}
\begin{algorithmic}
    \STATE Initialize $e_{seq}$ to $e_{full}$, and initialize $q$ to 0.
    \WHILE{$q < K \cdot (L_Q - 1)$}
	    \FOR{the (at most) $K$ top layer packets}
		    \STATE Find the packet $m$ which has the least RD impact $\Phi_m$
		\ENDFOR
		\STATE Set the priority value of packet $m$ to $q$
		\STATE Remove $m$ from the sequence
		\STATE Update $e_{seq}$ to $e_{seq} + e_m$, and update $q$ to $q+1$
    \ENDWHILE
\end{algorithmic}
\end{algorithm}

It is easy to see that the computational complexity of this priority assignment algorithm is $O(K^2)$. Note that, the computation in this priority assignment process is actually very insignificant compared with the decoding process needed in the error vector acquisition. With a larger optimization window size $K$, we would have more packets to compare and determine their priorities each time, which will potentially result in more optimized extraction. However, doing this will also bring more computation and, more noticeably, increase the memory used to store the error vectors. Besides, if the priority values are to be stored and signalled using the 6-bit priority identifier field of the NALU header \cite{Amonou07}, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

In some scenarios, the priority assignment and bitstream extraction needs to be conducted without access to the original video sequence. This would make the bitstream extraction a more difficult problem, since the distortion is harder to obtain for lacking of reference signal. The ``no-reference bitstream extraction'' problem is actually more frequently encountered in real practice, since it is more common for the streaming service provider to have the compressed bitstream rather than the video source. In this subsection, we consider this no-reference case and modify the proposed algorithm to meet the special characteristic caused by lacking of reference.

Without the original video sequence, we cannot get the error between the video sequence decoded from full enhancement packets and the original sequence, which is $e_{full}$ used in \ref{algo:greedy}. A straightforward idea is to set $e_{full}$ to zero, meaning all distortions are calculated with respect to the fully reconstructed signal, rather than to the original sequence. However, simply doing so causes the algorithm to perform badly in experiments. It is found that, since discarding some enhancement packets will not bring much significant error, the calculated PSNR values are too large to well represent the real distortion. To address this issue, instead of calculating PSNR, we directly use MSE of all pixels as the distortion criteria in the priority assignment algorithm. This proves to be an effective solution, and the extraction performance turns out to be much better.

@@

PID controller \cite{PID}\cite{Astrom02} is a control loop feedback mechanism commonly used in industrial control systems. \textit{Process variable} and \textit{setpoint} are two core concepts in PID control. The process variable stands for the current status of the system. The setpoint is the target for this system to reach. The goal of PID control is to minimize the error between process variable and setpoint. The controller output $u(t)$ is defined as follows:
\begin{equation}
\label{eq:pid-output}
u(t) = {K_p} \cdot e(t) + {K_i} \cdot \int_0^t {e(\tau )d\tau }  + {K_d} \cdot \frac{d}{{dt}}e(t) \: ,
\end{equation}
where $e(t)$ represents the error between process variable and setpoint at time $t$. $K_p$, $K_i$, and $K_d$ here are tuning parameters representing the proportional gain, integral gain and derivative gain, respectively. The result of (\ref{eq:pid-output}) is used for manipulating the system to make process variable change towards setpoint.

As an effective and flexible control technique, PID control has been used to solve all kinds of control problems. In the area of video coding, we notice that PID is utilized to design rate control algorithms (rate control in video coding is mainly to adjust the encoding parameters for every frame, so that the bits are appropriately allocated and desired bitrate is achieved) \cite{Wong04}\cite{Yang10}. We also notice that some researchers have proposed theoretical model or middle-ware framework for QoS adaptation \cite{Li98}\cite{Li99}, with PID as the basic control scheme. However, we have not found any paper that integrates PID method into the quality control of video streaming. While rate control of video coding and quality control of video streaming are two different problems, they do share some common properties, such as the buffer mechanism and the requirement for smoothness. That inspires us to design a PID-based quality control algorithm and utilize it to deliver better video quality to the user.

@@

In DSS, two timelines are important. Push timeline indicates the time of pushing video data packets to streaming buffer, and play timeline indicates the time these data packet should be played, i.e., the presentation time stamp (often seen as PTS) of each packet. Their relationship is shown in Fig. \ref{fig:intervals}. The program will check the timelines periodically, and, in the default quality control algorithm of DSS, use the packet delays to decide whether quality change is needed. The two intervals shown in Fig. \ref{fig:intervals}, i.e., check interval and actual interval, are worth noticing and may provide useful information. In the ideal situation, during each check interval, the server pushes an amount of data that will actually play for a period of time equal to that check interval, i.e., \textit{actual\_interval} = \textit{check\_interval}. If \textit{actual\_interval} is smaller than \textit{check\_interval}, it means that less data is pushed to the client, and the pushing is slowed down because bandwidth is relatively low at this time. On the other hand, if \textit{actual\_interval} is larger than \textit{check\_interval}, it indicates the network condition is good and data is being pushed quickly to the client. The ratio of these two intervals, i.e., \textit{actual\_check interval ratio}, is in direct proportion to the ratio of the bitrate suitable for the network condition and the bitrate currently being used.

@@

This ratio-based model is more simple and effective for the chosen process variable and setpoint. Moreover, in this model the controller output has intuitive meaning when the three parameters are normalized. It is easy to know that $E_p^t$, $E_i^t$, $E_d^t$ are all ratios with a value more or less than 1, and the controller output $u_t$ can be seen as a weighted average of the three ratios, with the three parameters as weights. Therefore the controller output is also a ratio value around 1, which reflects the relationship between the bitrate suitable for the network condition and the bitrate currently being used. In other words, if the output is larger than 1, then the controller has determined the bitrate should be increased accordingly, and vice versa. Use it as a multiplier to manipulate the bitrate currently being used, and we would be able to control the streaming system to match the network condition.
